---
title: "Aus23_Microbial_soil_data.Rmd"
output: html_document
date: "2024-10-17"
---

This is the code to merge the ITS and 16S data summary statistics with the soil nutrient data.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

load libraries
```{r}
library(tidyverse)
library(stringr)
```


Read in all the dataframes
```{r}
###soil###
CN <- read.csv("../Soil_data/Aus_IRES_totCN.csv")
erg_P <- read.csv("../Soil_data/Aus_IRES_ergosterol_P.csv")

###microbial###

#bacterial guild abundances and diversity
bac_metrics <- read.csv("../Aus23_16S_Metabarcoding/OTUtables/AlphaDivGuilds16S.csv", row.names=1)
bac_metrics_pool <- read.csv("../Aus23_16S_Metabarcoding/OTUtables/AlphaDivGuilds16S_pooled.csv", row.names=1)

#fungal guild abundances and diversity
fungal_metrics <- read.csv("../Aus23_ITS_Metabarcoding/OTUTables/AlphaDivGuildsITS.csv", row.names=1)
fungal_metrics_pool <- read.csv("../Aus23_ITS_Metabarcoding/OTUTables/AlphaDivGuildsITS_pooled.csv", row.names=1)

#AMF diversity
amf_metrics <- read.csv("../Aus23_18S_Metabarcoding/OTUTables/AlphaDiv18S.csv", row.names=1)
amf_metrics_pool <- read.csv("../Aus23_18S_Metabarcoding/OTUTables/AlphaDiv18S_pooled.csv", row.names=1)

```

Create a pooled version of the total CN dataset that averages measurements by tree ID
```{r}
# create a unique tree ID
CN$treeName <- str_sub(CN$Sample, 1, 2)
# average the two reps from each tree
CN_pooled <- aggregate(cbind(perc_N, perc_C, CN_ratio) ~ treeName, data = CN, FUN = mean)

```


Format column names and remove duplicate columns across dataframes
```{r}
### rename columns
colnames(bac_metrics)[1:4] <- c("Bac_Shannon", "Bac_Simpson", "Bac_Invsimpson", "Bac_Richness")
colnames(bac_metrics_pool)[1:4] <- c("Bac_Shannon", "Bac_Simpson", "Bac_Invsimpson", "Bac_Richness")
colnames(fungal_metrics)[1:4] <- c("Fungal_Shannon", "Fungal_Simpson", "Fungal_Invsimpson", "Fungal_Richness")
colnames(fungal_metrics_pool)[1:4] <- c("Fungal_Shannon", "Fungal_Simpson", "Fungal_Invsimpson", "Fungal_Richness")
colnames(amf_metrics)[1:4] <- c("AMF_Shannon", "AMF_Simpson", "AMF_Invsimpson", "AMF_Richness")
colnames(amf_metrics_pool)[1:4] <- c("AMF_Shannon", "AMF_Simpson", "AMF_Invsimpson", "AMF_Richness")

### remove duplicate columns 
#unpooled
fungal_metrics_sub <- fungal_metrics[c(1:5,12:23)]
amf_metrics_sub <- amf_metrics[1:5]
#pooled
fungal_pool_sub <- fungal_metrics_pool[c(1:5, 12:23)]
amf_pool_sub <- amf_metrics_pool[c(1:5)]

```


Merge all microbial datasets
```{r}
all_microbe <- bac_metrics %>%
  full_join(fungal_metrics_sub, by = "SampleID") %>%
  full_join(amf_metrics_sub, by = "SampleID")
# save microbial dataset
write.csv(all_microbe, "../Merged_data/Aus23_All_microbial_metrics_unpooled.csv")

all_microbe_pooled <- bac_metrics_pool %>%
  full_join(fungal_pool_sub, by = "TreeSampleType") %>%
  full_join(amf_pool_sub, by = "TreeSampleType")
# save microbial dataset
write.csv(all_microbe, "../Merged_data/Aus23_All_microbial_metrics_unpooled.csv")
write.csv(all_microbe_pooled, "../Merged_data/Aus23_All_microbial_metrics_unpooled.csv")


```

Separate out root and soil data
```{r}
#unpooled
all_microbe_soil <- subset(all_microbe, all_microbe$SampleType =="Soil")

#pooled
all_microbe_pool_soil <- subset(all_microbe_pooled, all_microbe_pooled$SampleType =="Soil")
all_microbe_pool_root <- subset(all_microbe_pooled, all_microbe_pooled$SampleType =="Root")


```


Merge unpooled soil microbial data with total CN measurements.
```{r}
# format sample names to be the same in both datasets
all_microbe_soil$Sample <- str_c(str_sub(all_microbe_soil$SampleID, 1, 2), str_sub(all_microbe_soil$SampleID, 4, 4))
#merge dataframes
microbes_soilCN <- merge(all_microbe_soil, CN, by = "Sample")
#save dataframe
write.csv(microbes_soilCN, "../Merged_data/Aus23_microbes_soilCN_unpooled.csv")

```

Merge pooled soil microbial data with total P, ergosterol, and total CN measurements. We use the pooled data for root DNA sequences, ergosterol, and total P because root DNA, total P, and ergosterol were measured on soils pooled across the N and S sampling points for each tree.
```{r}
# format sample names to be the same in all datasets
all_microbe_pool_soil$treeName <- str_sub(all_microbe_pool_soil$SampleID, 1, 2)
all_microbe_pool_root$treeName <- str_sub(all_microbe_pool_root$SampleID, 1, 2)

#remove duplicate columns between soil and root dataframes
microbe_pool_root_sub <- all_microbe_pool_root[c(1:4, 12:38)]

# because we are merging soil and root data, append soil metric column names with "_soil" and root metrics with "_root"
colnames(microbe_pool_root_sub)[1:30] <- paste0(colnames(microbe_pool_root_sub)[1:30], "_root")
colnames(all_microbe_pool_soil)[c(1:4, 12:37)] <- paste0(colnames(all_microbe_pool_soil)[c(1:4, 12:37)], "_soil")

#merge soil and root microbial data with soil nutrient data
microbe_soil_pooled <- all_microbe_pool_soil %>%
  full_join(microbe_pool_root_sub, by = "treeName") %>%
  full_join(CN_pooled, by = "treeName") %>%
  full_join(erg_P, by = "treeName")

#remove unneccesary columns
microbe_soil_pooled <- microbe_soil_pooled[c(1:4, 7, 12:74)]

#save as csv
write.csv(microbe_soil_pooled, "../Merged_data/Aus23_microbes_soilCNP_pooled.csv")

```


