---
title: "Aus23_merge_litter_data.Rmd"
output: html_document
date: "2024-10-17"
---

This is the code to merge the master dataframe with the litter data.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

load libraries
```{r}
library(tidyverse)
library(stringr)
library(DescTools)
```

Read in dataframes
```{r}
main <- read.csv("../Merged_data/Aus23_master_pooled.csv")
lit_erg <- read.csv("../Litter_data/Belanglo_litter_ergosterol.csv")
lit_p <- read.csv("../Litter_data/Belanglo_litter_P.csv")
lit_cn <- read.csv("../Litter_data/Belanglo_litter_CN.csv")

#select out columns we want
lit_erg <- lit_erg[c(1,4)]
lit_p <- lit_p[c(1,9,10)]
lit_cn <- lit_cn[c(2:4)]
```

Right now the master pooled datasheet has repeated data because each biolog plate day is in the same column. Separate out the biolog reading for each substrate by day in new columns (ie, each substrate has 3 columns for day 1, day 3 and day 5)
```{r}
# remove duplicate rows for the non-biolog data
main_nobio <- subset(main, main$day == 1)
main_nobio <- main_nobio[c(2,3, 37,39:119)]

# pull out just biolog data
biolog <- main[c(4:37)]

# separate by day
biolog1 <- subset(biolog, biolog$day ==1)
biolog3 <- subset(biolog, biolog$day ==3)
biolog5 <- subset(biolog, biolog$day ==5)
# edit column names
colnames(biolog1)[2:33] <- paste0(colnames(biolog1)[2:33], "_BiologDay1")
colnames(biolog3)[2:33] <- paste0(colnames(biolog3)[2:33], "_BiologDay3")
colnames(biolog5)[2:33] <- paste0(colnames(biolog5)[2:33], "_BiologDay5")
# join back together
biolog_wide <- biolog1[2:34] %>%
  full_join(biolog3[2:34], by = "treeName") %>%
  full_join(biolog5[2:34], by = "treeName") 
  
# merge back with main dataframe
main_edited <- merge(main_nobio, biolog_wide, by = "treeName")

```

Merge all litter data into one dataframe
```{r}
all_lit <- lit_erg %>%
  full_join(lit_p, by = "ID") %>%
  full_join(lit_cn, by = "ID")
```

Create new columns for tree IDs and litter types
```{r}
# paste tree name
all_lit$treeName <- str_sub(all_lit$ID, 1, 2)

# paste litter tree origin (pine or eucalyptus)
all_lit <- all_lit %>%
  mutate(litter_species = case_when(
    grepl("Pine", ID) ~ "Pine",
    grepl("Euc", ID) ~ "Euc",
    TRUE ~ NA_character_
  ))

# paste litter type (leaf litter, bark, or organic layer)
all_lit <- all_lit %>%
  mutate(litter_type = case_when(
    grepl("LL", ID) ~ "Leaf litter",
    grepl("B", ID) ~ "Bark",
    grepl("OL", ID) ~ "Organic layer",
    TRUE ~ NA_character_
  ))

# Create new dataframes by litter type
leaf.pine <- subset(all_lit, all_lit$litter_type %like% "Leaf litter" & all_lit$litter_species %like% "Pine")
leaf.euc <- subset(all_lit, all_lit$litter_type %like% "Leaf litter" & all_lit$litter_species %like% "Euc")
bark <- subset(all_lit, all_lit$litter_type %like% "Bark")
OL <- subset(all_lit, all_lit$litter_type %like% "Organic layer")

```

Subset out columns and rename columns
```{r}
leaf.euc <- leaf.euc[c(2:7)]
colnames(leaf.euc)[1:5] <- c("Euc_leafLitter_ergosterol", "Euc_leafLitter_totP", "Euc_leafLitter_percP", "Euc_leafLitter_percN", "Euc_leafLitter_percC")

leaf.pine <- leaf.pine[c(2:7)]
colnames(leaf.pine)[1:5] <- c("Pine_leafLitter_ergosterol", "Pine_leafLitter_totP", "Pine_leafLitter_percP", "Pine_leafLitter_percN", "Pine_leafLitter_percC")

bark <- bark[c(2:7)]
colnames(bark)[1:5] <- c("Euc_Bark_ergosterol", "Euc_Bark_totP", "Euc_Bark_percP", "Euc_Bark_percN", "Euc_Bark_percC")

OL <- OL[c(2:7)]
colnames(OL)[1:5] <- c("Litter_OLayer_ergosterol", "Litter_OLayer_totP", "Litter_OLayer_percP", "Litter_OLayer_percN", "Litter_OLayer_percC")
#remove the one row with tree name "ER"...not sure what this row is??
OL <- OL[c(1:18),]
```

Merge the formatted litter data into one dataframe
```{r}
lit_format <- leaf.euc %>%
  full_join(leaf.pine, by = "treeName") %>%
  full_join(bark, by = "treeName") %>%
  full_join(OL, by = "treeName")
```

Create average litter nutrient content across litter types
```{r}
lit_format$Litter_avg_ergosterol <- rowMeans(lit_format[, c('Euc_leafLitter_ergosterol', 'Pine_leafLitter_ergosterol','Euc_Bark_ergosterol', 'Litter_OLayer_ergosterol')], na.rm = TRUE)
lit_format$Litter_avg_totP <- rowMeans(lit_format[, c('Euc_leafLitter_totP', 'Pine_leafLitter_totP','Euc_Bark_totP', 'Litter_OLayer_totP')], na.rm = TRUE)
lit_format$Litter_avg_percP <- rowMeans(lit_format[, c('Euc_leafLitter_percP', 'Pine_leafLitter_percP','Euc_Bark_percP', 'Litter_OLayer_percP')], na.rm = TRUE)
lit_format$Litter_avg_percN <- rowMeans(lit_format[, c('Euc_leafLitter_percN', 'Pine_leafLitter_percN','Euc_Bark_percN', 'Litter_OLayer_percN')], na.rm = TRUE)
lit_format$Litter_avg_percC <- rowMeans(lit_format[, c('Euc_leafLitter_percC', 'Pine_leafLitter_percC','Euc_Bark_percC', 'Litter_OLayer_percC')], na.rm = TRUE)

# save final litter spreadsheet
write.csv(lit_format, "../Litter_data/Aus23_litter_allNutrients.csv")
  
```

Merge all into one final dataframe
```{r}
final <- merge(main_edited, lit_format, by = "treeName")
write.csv(final, "../Merged_data/Aus23_allData_7Nov24.csv")
```

