---
title: "ITS_DataImport_AlphaBetaDiv"
author: "Jake Nash"
date: "2024-04-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load Required Packages
```{r, warning=FALSE, message=FALSE}
library(tidyverse)
library(vegan)
library(car)
library(magrittr)
library(ecodist)
library(ANCOMBC)
library(phyloseq)
library(readxl)
```

## Import taxonomy file and parse taxonomy string into separate columns
```{r, warning=FALSE, message=FALSE}
extract_text <- function(x) {
  sub(".+__", "", x)
}

TaxMatITS=
  read_tsv("../Aus23_ITS_Metabarcoding/ITS2_Dada2_repseqs97_taxonomy.tsv") %>%
  rename("otu"="Feature ID") %>%
  select(-Confidence) %>%
  separate_wider_delim(too_few="align_start", cols = Taxon, delim = ";", names = c("Kingdom", "Phylum","Class", "Order","Family","Genus","Species", "SH")) %>%
  mutate(across(!contains("otu"), ~extract_text(.)))
```

## Import ASV table
```{r, warning=FALSE, message=FALSE}
OtuMatITS=
  read_tsv("../Aus23_ITS_Metabarcoding/ITS2_OTUTable_97.tsv", skip=1) %>%
  column_to_rownames("#OTU ID")
```

## Import Metadata
```{r, warning=FALSE, message=FALSE}
Metadata_ITS=
  read_tsv("../Aus23_16S_Metabarcoding/MetabarcodingMetadata.txt") %>%
  slice(-1) %>%
  rename("SampleID"=`sample-id`)
```

## Calculates sequencing depth per sample, then rarefies to 70218
Rarefaction curves are generated showing rarefaction point as dotted line. Rarefaction resulted in dropping sample P3-2_Root and the negative control. OTUs that had 0 sequences remaining after rarefaction were removed. Random seed was set to ensure reproducibility
```{r}
SeqDepthITS=
  OtuMatITS %>%
  colSums() %>%
  as.data.frame %>%
  set_colnames("n_sequences")

RareCurve_ITs=
  OtuMatITS %>%
  t %>%
  rarecurve(step=500, tidy=TRUE)

ggplot(RareCurve_ITs, aes(x=Sample, y=Species, group=Site)) +
  geom_line() +
  geom_vline(xintercept=70218, linetype=2) +
  ylab("# ASVs") +
  xlab("# Sequences")

set.seed(2023)
OtuMatITS_rare=
  OtuMatITS %>%
  t %>%
  as.data.frame %>%
  rrarefy(sample=70218) %>%
  as.data.frame %>%
  select(-which(colSums(.) == 0)) %>%
  filter(rowSums(.) >= 70218)
```

## Conducts PCoA with Aitchison distance
PCoA is first plotted with points colored by sample type. Then PCoA is plotted colored by tree species, this time using axes 2 and 3 because axis 3 shows separation by tree species best.
```{r, warning=FALSE, message=FALSE}
DistMatITS=vegdist(OtuMatITS_rare, method = "aitchison", pseudocount=1)
PCoA_ITS_aitchison<- ecodist::pco(DistMatITS)
PCoA_ITS_aitchison_df=
  PCoA_ITS_aitchison$vectors %>%
  as.data.frame %>%
  cbind(data.frame(SampleID=rownames(.)), .) %>%
  left_join(Metadata_ITS)

ggplot(PCoA_ITS_aitchison_df, aes(x=X1, y=X2)) +
  geom_point(aes(color=SampleType))

PCoA_ITS_aitchison_df %>%
  ggplot(aes(x=X1, y=X2)) +
  geom_point(aes(color=TreeSpecies, shape=SampleType))
```

## Conducts PERMANOVA with SampleType*TreeSpecies
```{r, warning=FALSE, message=FALSE}
ITS_PERMANOVA_SampleType=adonis2(DistMatITS~SampleType*TreeSpecies,data=PCoA_ITS_aitchison_df)
ITS_PERMANOVA_SampleType
```

## Generates alpha diversity dataframe, plots, then ANOVA
```{r, warning=FALSE, message=FALSE}
AlphaDiv_ITS=
  data.frame(Shannon=diversity(OtuMatITS_rare, index="shannon"),
             Simpson=diversity(OtuMatITS_rare, index="simpson"),
             Invsimpson=diversity(OtuMatITS_rare, index="invsimpson"),
             Richness=specnumber(OtuMatITS_rare)) %>%
  mutate(SampleID=rownames(.)) %>%
  left_join(Metadata_ITS)

ggplot(AlphaDiv_ITS, aes(x=SampleType, y=Richness, color=TreeSpecies)) +
  geom_boxplot()

ggplot(AlphaDiv_ITS, aes(x=SampleType, y=Shannon, color=TreeSpecies)) +
  geom_boxplot()

AlphaDiv_ITS %>%
  lm(Richness~SampleType*TreeSpecies,.) %>%
  Anova

AlphaDiv_ITS %>%
  lm(Shannon~SampleType*TreeSpecies,.) %>%
  Anova
```


## Conducts differential abundance testing with ANCOM-BC
Data is first converted to a phyloseq object because ANCOMBC needs it in that format. The dataset is divided into root samples and soil samples and separate tests are run for each subset to look for effect of tree species 
```{r, warning=FALSE, message=FALSE}
OTU_ITS_phyloseq=
  OtuMatITS_rare %>%
  as.matrix %>%
  otu_table(taxa_are_rows = FALSE)

TAX_ITS_phyloseq=
  TaxMatITS %>%
  filter(otu %in% colnames(OtuMatITS_rare)) %>%
  column_to_rownames("otu") %>%
  as.matrix %>%
  tax_table
  
SAMPLES_ITS_phyloseq=
  Metadata_ITS %>%
  filter(SampleID %in% rownames(OTU_ITS_phyloseq)) %>%
  column_to_rownames("SampleID") %>%
  sample_data(Metadata_16S)

PhyloseqITS <- phyloseq(OTU_ITS_phyloseq, TAX_ITS_phyloseq, SAMPLES_ITS_phyloseq)

PhyloseqITS_root=subset_samples(PhyloseqITS, SampleType=="Root")
ANCOM_ITS_root = ancombc(data = PhyloseqITS_root, assay_name = "counts", 
              tax_level = "Species", phyloseq = NULL, 
              formula = "TreeSpecies", 
              group="TreeSpecies", p_adj_method = "holm", prv_cut = 0.10, lib_cut = 1000, 
              struc_zero = TRUE, neg_lb = TRUE, tol = 1e-5, 
              max_iter = 100, conserve = TRUE, alpha = 0.05, global = TRUE,
              n_cl = 1, verbose = TRUE)

p_thresh=0.05

ANCOM_ITS_root_sig= 
  ANCOM_ITS_root$res$p_val %>% 
  as.data.frame %>% 
  filter(TreeSpeciesPine<p_thresh) %$%
  taxon

ANCOM_ITS_root_results=
  ANCOM_ITS_root$res$lfc %>% 
  as.data.frame %>% 
  filter(taxon %in% ANCOM_ITS_root_sig) %>%
  rename("LFC"="TreeSpeciesPine") %>%
  mutate(Enrichment=ifelse(LFC<0, "Eucalyptus_enriched", "Pine_enriched"))

PhyloseqITS_soil=subset_samples(PhyloseqITS, SampleType=="Soil")
ANCOM_ITS_soil = ancombc(data = PhyloseqITS_soil, assay_name = "counts", 
              tax_level = "Species", phyloseq = NULL, 
              formula = "TreeSpecies", 
              group="TreeSpecies", p_adj_method = "holm", prv_cut = 0.10, lib_cut = 1000, 
              struc_zero = TRUE, neg_lb = TRUE, tol = 1e-5, 
              max_iter = 100, conserve = TRUE, alpha = 0.05, global = TRUE,
              n_cl = 1, verbose = TRUE)

ANCOM_ITS_soil_sig= 
  ANCOM_ITS_soil$res$p_val %>% 
  as.data.frame %>% 
  filter(TreeSpeciesPine<p_thresh) %$%
  taxon

ANCOM_ITS_soil_results=
  ANCOM_ITS_soil$res$lfc %>% 
  as.data.frame %>% 
  filter(taxon %in% ANCOM_ITS_soil_sig) %>%
  rename("LFC"="TreeSpeciesPine") %>%
  mutate(Enrichment=ifelse(LFC<0, "Eucalyptus_enriched", "Pine_enriched"))

ANCOM_ITS_root_results %>%
  arrange(LFC) %>%
  slice(c(1:20, (n() - 19):n())) %>%
  arrange(desc(LFC)) %>%
  ggplot(aes(x=reorder(taxon, LFC), y=LFC, fill=Enrichment)) +
  geom_bar(stat="identity") +
  coord_flip() +
  theme_minimal() +
  ggtitle("Plant Roots - Eucalyptus vs. Pine - Top 40 enriched taxa") +
  geom_hline(yintercept=0)


ANCOM_ITS_soil_results %>%
  arrange(LFC) %>%
  slice(c(1:20, (n() - 19):n())) %>%
  arrange(desc(LFC)) %>%
  ggplot(aes(x=reorder(taxon, LFC), y=LFC, fill=Enrichment)) +
  geom_bar(stat="identity") +
  coord_flip() +
  theme_minimal() +
  ggtitle("Soils - Eucalyptus vs. Pine - Top 40 enriched taxa") +
  geom_hline(yintercept=0)

```


## Plots total abundance of all Suillus OTUs
```{r, warning=FALSE, message=FALSE}
SuillOTUS=
  TaxMatITS %>%
  filter(Genus=="Suillus", .$otu %in% colnames(OtuMatITS_rare)) %$%
  otu

SuillusAbundance=
  OtuMatITS_rare %>%
  select(SuillOTUS) %>%
  mutate(Suillus=rowSums(.)) %>%
  select(Suillus) %>%
  mutate(SampleID=rownames(.)) %>%
  left_join(Metadata_ITS)

ggplot(SuillusAbundance, aes(x=SampleType, y=log(Suillus+1), color=TreeSpecies)) +
  geom_boxplot() +
  geom_point(position=position_dodge(width=0.75),aes(group=TreeSpecies))

```

## Plots abundance of each Suillus OTU separately
```{r, warning=FALSE, message=FALSE}
UpdatedTax=read_excel("../Aus23_ITS_Metabarcoding/ITS2_Dada2_repseqs97_taxonomy_edited.xlsx")

Suillus_OTUs_Abundance=
  OtuMatITS_rare %>%
  select(SuillOTUS) %>%
  mutate(SampleID=rownames(.)) %>%
  left_join(Metadata_ITS) %>%
  gather(otu, Abundance, -TreeSpecies, -SampleType, -TreeID, -Plot_Rep, -SampleID, -TreeSpecies_SampleType) %>%
  left_join(UpdatedTax,by=c("otu"="Feature ID")) %>%
  mutate(SpeciesOTU=paste(Species, otu, sep="_"))

ggplot(Suillus_OTUs_Abundance, aes(x=SampleType, y=log(Abundance+0.01), color=TreeSpecies)) +
  geom_boxplot() +
  geom_point(position=position_dodge(width=0.75),aes(group=TreeSpecies)) +
  facet_wrap(.~SpeciesOTU) +
  theme(strip.text.x = element_text(angle = 0, hjust = 0))


```
