---
title: "LinDA_DifferentialAbundance_18S"
output: html_document
date: "2024-10-15"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require("knitr")
opts_knit$set(root.dir = "/Users/alex/Documents/Documents - Alex’s MacBook Pro/Documents/GitHub/Aus-Invasions-2023-Course/Aus23_18S_Metabarcoding")
```

These scripts use LinDA to perform differential abundance testing to test for effects of tree species on OTU abundance. Tests are performed separately for root and soil samples.

The LinDA method was done both with and without the random effect of Tree individual to test how it affected statistical power.

#Load required packages
*Note that the MicrobiomeStat package contains the code for LINDA
```{r}
library(MicrobiomeStat)
library(dplyr)
library(magrittr)
library(tibble)
library(ANCOMBC)
library(tidyverse)
library(readxl)
```

#Reads in raw 18S OTU tables & metadata
```{r}
OtuMat18S = read_tsv("AMF_FeatureTable.tsv") %>%
  column_to_rownames("#OTU ID")

Metadata_ITS=
  read_tsv("../Aus23_16S_Metabarcoding/MetabarcodingMetadata.txt") %>%
  slice(-1) %>% #remove the comment row 
  rename("SampleID"=`sample-id`) %>% #rename SampleID column
  mutate(TreeSampleType = gsub("-.", "", SampleID)) #generate new column with Tree ID and sample type concatenated

TaxMat18S = read.csv("AMF_Taxonomy_SplitFormat.csv") %>%
  rename("otu"="Feature.ID")
```

#LinDA test for 18S root samples
```{r}
#create ITS OTU table just with Root samples
OtuMat18S_root= 
  OtuMat18S%>% 
  select(contains("Root")) 

#create metadata table just with root samples in ITS otu table
Metadata_18S_root=
  Metadata_18S %>%
  filter(SampleID %in% colnames(OtuMat18S_root))

#runs LinDA test for effect of tree species, using tree individual as random effects
Linda_18S_root =
  linda(
    feature.dat=OtuMat18S_root,
    meta.dat=Metadata_18S_root,
    formula='~Species + (1|TreeSampleType)',
    feature.dat.type="count",
    prev.filter=0.1,
    zero.handling="pseudo.cnt",
    pseudo.cnt = 0.5,
    p.adj.method="fdr",
    alpha=0.05)

#runs LinDA test for effect of tree species with no random effect
Linda_18S_root_no_random =
  linda(
    feature.dat=OtuMat18S_root,
    meta.dat=Metadata_18S_root,
    formula='~Species',
    feature.dat.type="count",
    prev.filter=0.1,
    zero.handling="pseudo.cnt",
    pseudo.cnt = 0.5,
    p.adj.method="fdr",
    alpha=0.05)

#creates a formatted dataframe for 18S root samples LinDA
Linda_18S_root_df=
  as.data.frame(Linda_18S_root$output$SpeciesPine) %>%
  mutate(otu=rownames(.)) %>%
  left_join(TaxMat18S)

Linda_18S_root_no_random_df=
  as.data.frame(Linda_18S_root_no_random$output$SpeciesPine) %>%
    mutate(otu=rownames(.)) %>%
  left_join(TaxMat18S)
```

#LinDA test on 18S soil samples
```{r}
#create 18S OTU table just with Soil samples
OtuMat18S_soil= 
  OtuMat18S %>% 
  select(contains("Soil")) 
  
#create metadata table just with soil samples in 18S otu table
Metadata_18S_soil=
  Metadata_ITS %>%
  filter(SampleID %in% colnames(OtuMat18S_soil))

#runs LinDA test for effect of tree species, using tree individual as random effects
Linda_18S_soil =
  linda(
    feature.dat=OtuMat18S_soil,
    meta.dat=Metadata_18S_soil,
    formula='~TreeSpecies + (1|TreeSampleType)',
    feature.dat.type="count",
    prev.filter=0.1,
    zero.handling="pseudo.cnt",
    pseudo.cnt = 0.5,
    p.adj.method="fdr",
    alpha=0.05)

#runs LinDA test for effect of tree species with no random effect
Linda_18S_soil_no_random =
  linda(
    feature.dat=OtuMat18S_soil,
    meta.dat=Metadata_18S_soil,
    formula='~TreeSpecies',
    feature.dat.type="count",
    prev.filter=0.1,
    zero.handling="pseudo.cnt",
    pseudo.cnt = 0.5,
    p.adj.method="fdr",
    alpha=0.05)

#creates a formatted dataframe for ITS root samples LinDA
Linda_18S_soil_df=
  as.data.frame(Linda_18S_soil$output$TreeSpeciesPine) %>%
  mutate(otu=rownames(.)) %>%
  left_join(TaxMat18S)

Linda_18S_soil_no_random_df=
  as.data.frame(Linda_18S_soil_no_random$output$TreeSpeciesPine) %>%
  mutate(otu=rownames(.)) %>%
  left_join(TaxMat18S)
```

#Create table of differentially abundant taxa
```{r}
n_18S_root=
  Linda_18S_root_df %>%
  filter(reject) %>%
  nrow

n_18S_soil=
  Linda_18S_soil_df %>%
  filter(reject) %>%
  nrow

n_18S_root_no_random=
  Linda_18S_root_no_random_df %>%
  filter(reject) %>%
  nrow

n_18S_soil_no_random=
  Linda_18S_soil_no_random_df %>%
  filter(reject) %>%
  nrow

LINDA_summary=
  data.frame(Locus=c("18S", "18S", "18S", "18S"),
             SampleType=c("Root","Soil","Root","Soil"),
             ModelType=c("RandomEffect","RandomEffect","NoRandomEffect","NoRandomEffect"),
             n_taxa_significant=c(n_18S_root, n_18S_soil,
                                  n_18S_root_no_random, n_18S_soil_no_random))

LINDA_summary

```

