---
title: "LINDA Differential Abundance Testings"
author: "Jake Nash"
date: "2024-10-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

These scripts use LinDA to perform differential abundance testing to test for effects of tree species on OTU abundance. Tests are performed separately for root and soil samples. 

The LinDA method was done both with and without the random effect of Tree individual to test how it affected statistical power.

## Load required packages
Note that the MicrobiomeStat package contains the code for LINDA
```{r, warning=FALSE, message=FALSE}
library(MicrobiomeStat)
library(dplyr)
library(magrittr)
library(tibble)
library(ANCOMBC)
library(tidyverse)
library(readxl)
```

## Reads in raw ITS OTU tables and metadata

```{r, warning=FALSE, message=FALSE}
OtuMatITS=
  read_tsv("../Aus23_ITS_Metabarcoding/ITS2_OTUTable_97.tsv", skip=1) %>%
  column_to_rownames("#OTU ID")

Metadata_ITS=
  read_tsv("../Aus23_16S_Metabarcoding/MetabarcodingMetadata.txt") %>%
  slice(-1) %>% #remove the comment row 
  rename("SampleID"=`sample-id`) %>% #rename SampleID column
  mutate(TreeSampleType = gsub("-.", "", SampleID)) #generate new column with Tree ID and sample type concatenated

TaxMatITS=
  read_xlsx("../Aus23_ITS_Metabarcoding/ITS2_Dada2_repseqs97_taxonomy_edited.xlsx") %>%
  rename("otu"="Feature ID") %>%
  select(otu, Kingdom:Species)

OtuMat16S_filter=
  read.csv("../Aus23_16S_Metabarcoding/OTUTables/OtuMat16S_filter.csv", row.names = 1) %>%
  mutate(across(everything())) %>%
  rename_with(~ gsub("\\.", "-", .))

Metadata_16S=
  read_tsv("../Aus23_16S_Metabarcoding/MetabarcodingMetadata.txt") %>%
  slice(-1) %>% #remove the comment row 
  rename("SampleID"=`sample-id`) %>% #rename SampleID column
  mutate(TreeSampleType = gsub("-.", "", SampleID)) #generate new column with Tree ID and sample type concatenated

#custom function to parse taxonomy
extract_text <- function(x) {
  sub(".+__", "", x)
}

TaxMat16S_SILVA=
  read_tsv("../Aus23_16S_Metabarcoding/Aus23_16S_SilvaTaxonomy_16S.tsv") %>% 
  rename("otu"="Feature ID") %>%
  select(-Confidence) %>%
  separate_wider_delim(too_few="align_start", cols = Taxon, delim = ";", names = c("Kingdom", "Phylum","Class", "Order","Family","Genus","Species", "SH")) %>%
  mutate(across(!contains("otu"), ~extract_text(.)))
```

## LINDA test for ITS root samples
```{r, warning=FALSE, message=FALSE}
#create ITS OTU table just with Root samples
OtuMatITS_root= 
  OtuMatITS%>% 
  select(contains("Root")) 

#create metadata table just with root samples in ITS otu table
Metadata_ITS_root=
  Metadata_ITS %>%
  filter(SampleID %in% colnames(OtuMatITS_root))

#runs LinDA test for effect of tree species, using tree individual as random effects
Linda_ITS_root =
  linda(
    feature.dat=OtuMatITS_root,
    meta.dat=Metadata_ITS_root,
    formula='~TreeSpecies + (1|TreeSampleType)',
    feature.dat.type="count",
    prev.filter=0.1,
    zero.handling="pseudo.cnt",
    pseudo.cnt = 0.5,
    p.adj.method="fdr",
    alpha=0.05)

Linda_ITS_root_no_random =
  linda(
    feature.dat=OtuMatITS_root,
    meta.dat=Metadata_ITS_root,
    formula='~TreeSpecies',
    feature.dat.type="count",
    prev.filter=0.1,
    zero.handling="pseudo.cnt",
    pseudo.cnt = 0.5,
    p.adj.method="fdr",
    alpha=0.05)

#creates a formatted dataframe for ITS root samples LinDA
Linda_ITS_root_df=
  as.data.frame(Linda_ITS_root$output$TreeSpeciesPine) %>%
  mutate(otu=rownames(.)) %>%
  left_join(TaxMatITS)

Linda_ITS_root_no_random_df=
  as.data.frame(Linda_ITS_root_no_random$output$TreeSpeciesPine) %>%
    mutate(otu=rownames(.)) %>%
  left_join(TaxMatITS)
```

## LINDA test on ITS Soil samples
```{r, warning=FALSE, message=FALSE}
#create ITS OTU table just with Soil samples
OtuMatITS_soil= 
  OtuMatITS %>% 
  select(contains("Soil")) 
  
#create metadata table just with root samples in ITS otu table
Metadata_ITS_soil=
  Metadata_ITS %>%
  filter(SampleID %in% colnames(OtuMatITS_soil))

#runs LinDA test for effect of tree species, using tree individual as random effects
Linda_ITS_soil =
  linda(
    feature.dat=OtuMatITS_soil,
    meta.dat=Metadata_ITS_soil,
    formula='~TreeSpecies + (1|TreeSampleType)',
    feature.dat.type="count",
    prev.filter=0.1,
    zero.handling="pseudo.cnt",
    pseudo.cnt = 0.5,
    p.adj.method="fdr",
    alpha=0.05)

Linda_ITS_soil_no_random =
  linda(
    feature.dat=OtuMatITS_soil,
    meta.dat=Metadata_ITS_soil,
    formula='~TreeSpecies',
    feature.dat.type="count",
    prev.filter=0.1,
    zero.handling="pseudo.cnt",
    pseudo.cnt = 0.5,
    p.adj.method="fdr",
    alpha=0.05)

#creates a formatted dataframe for ITS root samples LinDA
Linda_ITS_soil_df=
  as.data.frame(Linda_ITS_soil$output$TreeSpeciesPine) %>%
  mutate(otu=rownames(.)) %>%
  left_join(TaxMatITS)

Linda_ITS_soil_no_random_df=
  as.data.frame(Linda_ITS_soil_no_random$output$TreeSpeciesPine) %>%
  mutate(otu=rownames(.)) %>%
  left_join(TaxMatITS)

```

## LINDA test for 16S root samples
```{r, warning=FALSE, message=FALSE}
#create 16S OTU table just with Root samples
OtuMat16S_root= 
  OtuMat16S_filter %>% 
  select(contains("Root")) 

#create metadata table just with root samples in 16S otu table
Metadata_16S_root=
  Metadata_16S %>%
  filter(SampleID %in% colnames(OtuMat16S_root))

#runs LinDA test for effect of tree species, using tree individual as random effects
Linda_16S_root =
  linda(
    feature.dat=OtuMat16S_root,
    meta.dat=Metadata_16S_root,
    formula='~TreeSpecies + (1|TreeSampleType)',
    feature.dat.type="count",
    prev.filter=0.1,
    zero.handling="pseudo.cnt",
    pseudo.cnt = 0.5,
    p.adj.method="fdr",
    alpha=0.05)

Linda_16S_root_no_random =
  linda(
    feature.dat=OtuMat16S_root,
    meta.dat=Metadata_16S_root,
    formula='~TreeSpecies',
    feature.dat.type="count",
    prev.filter=0.1,
    zero.handling="pseudo.cnt",
    pseudo.cnt = 0.5,
    p.adj.method="fdr",
    alpha=0.05)

#creates a formatted dataframe for 16S root samples LinDA
Linda_16S_root_df=
  as.data.frame(Linda_16S_root$output$TreeSpeciesPine) %>%
  mutate(otu=rownames(.)) %>%
  left_join(TaxMat16S_SILVA)

Linda_16S_root_no_random_df=
  as.data.frame(Linda_16S_root_no_random$output$TreeSpeciesPine) %>%
  mutate(otu=rownames(.)) %>%
  left_join(TaxMat16S_SILVA)
```

## LINDA test on 16S Soil samples
```{r, warning=FALSE, message=FALSE}
#create 16S OTU table just with Soil samples
OtuMat16S_soil= 
  OtuMat16S_filter %>% 
  select(contains("Soil")) 
  
#create metadata table just with root samples in 16S otu table
Metadata_16S_soil=
  Metadata_16S %>%
  filter(SampleID %in% colnames(OtuMat16S_soil))

#runs LinDA test for effect of tree species, using tree individual as random effects
Linda_16S_soil =
  linda(
    feature.dat=OtuMat16S_soil,
    meta.dat=Metadata_16S_soil,
    formula='~TreeSpecies + (1|TreeSampleType)',
    feature.dat.type="count",
    prev.filter=0.1,
    zero.handling="pseudo.cnt",
    pseudo.cnt = 0.5,
    p.adj.method="fdr",
    alpha=0.05)

Linda_16S_soil_no_random =
  linda(
    feature.dat=OtuMat16S_soil,
    meta.dat=Metadata_16S_soil,
    formula='~TreeSpecies',
    feature.dat.type="count",
    prev.filter=0.1,
    zero.handling="pseudo.cnt",
    pseudo.cnt = 0.5,
    p.adj.method="fdr",
    alpha=0.05)

#creates a formatted dataframe for 16S root samples LinDA
Linda_16S_soil_df=
  as.data.frame(Linda_16S_soil$output$TreeSpeciesPine) %>%
  mutate(otu=rownames(.)) %>%
  left_join(TaxMat16S_SILVA)

Linda_16S_soil_no_random_df=
  as.data.frame(Linda_16S_soil_no_random$output$TreeSpeciesPine) %>%
  mutate(otu=rownames(.)) %>%
  left_join(TaxMat16S_SILVA)

```

## Creates Phyloseq files for ANCOMBC-2
ANCOMBC-2 requires OTU table, taxonomy, and metadata be converted into a phyloseq format object. Phyloseq objects are also created with data subsetted into root and soil samples
```{r, warning=FALSE, message=FALSE}
OTU_ITS_phyloseq=
  OtuMatITS %>%
  as.matrix %>%
  otu_table(taxa_are_rows = TRUE)

TAX_ITS_phyloseq=
  TaxMatITS %>%
  filter(otu %in% rownames(OtuMatITS)) %>%
  column_to_rownames("otu") %>%
  as.matrix %>%
  tax_table

SAMPLES_ITS_phyloseq=
  Metadata_ITS %>% 
  filter(SampleID %in% colnames(OTU_ITS_phyloseq)) %>%
  column_to_rownames("SampleID") %>% 
  sample_data()

PhyloseqITS <- phyloseq(OTU_ITS_phyloseq, TAX_ITS_phyloseq, SAMPLES_ITS_phyloseq)

#subset phyloseq object into root and soil samples
PhyloseqITS_root=subset_samples(PhyloseqITS, SampleType=="Root")
PhyloseqITS_soil=subset_samples(PhyloseqITS, SampleType=="Soil")

OTU_16S_phyloseq=
  OtuMat16S_filter %>%
  as.matrix %>%
  otu_table(taxa_are_rows = TRUE)

TAX_16S_phyloseq=
  TaxMat16S_SILVA %>%
  filter(otu %in% rownames(OtuMat16S)) %>%
  column_to_rownames("otu") %>%
  as.matrix %>%
  tax_table

SAMPLES_16S_phyloseq=
  Metadata_16S %>%
  filter(SampleID %in% colnames(OTU_16S_phyloseq)) %>%
  column_to_rownames("SampleID") %>% 
  sample_data()

Phyloseq16S <- phyloseq(OTU_16S_phyloseq, TAX_16S_phyloseq, SAMPLES_16S_phyloseq)

#subsets 16S phyloseq object into root and soil
Phyloseq16S_root=subset_samples(Phyloseq16S, SampleType=="Root")
Phyloseq16S_soil=subset_samples(Phyloseq16S, SampleType=="Soil")
```

## ITS root ANCOMBC-2 test
```{r, warning=FALSE, message=FALSE}
ANCOM_ITS_root = ancombc2(data = PhyloseqITS_root, tax_level = NULL,
                  fix_formula = "TreeSpecies", rand_formula = "(1|TreeSampleType)",
                  p_adj_method = "fdr", pseudo_sens = TRUE,
                  prv_cut = 0.10, lib_cut = 1000, s0_perc = 0.05,
                  group = "TreeSpecies", struc_zero = FALSE, neg_lb = FALSE,
                  alpha = 0.05, n_cl = 2, verbose = TRUE,
                  global = TRUE, pairwise = FALSE, 
                  dunnet = FALSE, trend = FALSE,
                  iter_control = list(tol = 1e-5, max_iter = 20, 
                                      verbose = FALSE),
                  em_control = list(tol = 1e-5, max_iter = 100),
                  lme_control = lme4::lmerControl(), 
                  mdfdr_control = list(fwer_ctrl_method = "fdr", B = 100), 
                  trend_control = NULL)

ANCOM_ITS_root_no_random = ancombc2(data = PhyloseqITS_root, tax_level = NULL,
                  fix_formula = "TreeSpecies", rand_formula = NULL,
                  p_adj_method = "fdr", pseudo_sens = TRUE,
                  prv_cut = 0.10, lib_cut = 1000, s0_perc = 0.05,
                  group = "TreeSpecies", struc_zero = FALSE, neg_lb = FALSE,
                  alpha = 0.05, n_cl = 2, verbose = TRUE,
                  global = TRUE, pairwise = FALSE, 
                  dunnet = FALSE, trend = FALSE,
                  iter_control = list(tol = 1e-5, max_iter = 20, 
                                      verbose = FALSE),
                  em_control = list(tol = 1e-5, max_iter = 100),
                  lme_control = NULL, 
                  mdfdr_control = list(fwer_ctrl_method = "fdr", B = 100), 
                  trend_control = NULL)

ANCOM_ITS_root_df=
  ANCOM_ITS_root$res %>%
  as.data.frame %>% 
  filter(diff_TreeSpeciesPine) %>%
  rename("otu"="taxon") %>%
  left_join(TaxMatITS)

ANCOM_ITS_root_no_random_df=
  ANCOM_ITS_root_no_random$res %>%
  as.data.frame %>%
  filter(diff_TreeSpeciesPine) %>%
  rename("otu"="taxon") %>%
  left_join(TaxMatITS)
```

## Create Summary Data Table
Table just has counts of # OTUs differentially abundant in each test

Table shows there are VERY few OTUs differing by species in Linda tests. Just 1 significant OTU across the 4 tests
```{r, warning=FALSE, message=FALSE}

n_Linda_ITS_root=
  Linda_ITS_root_df %>%
  filter(reject) %>%
  nrow

n_Linda_ITS_soil=
  Linda_ITS_soil_df %>%
  filter(reject) %>%
  nrow

n_Linda_16S_root=
  Linda_16S_root_df %>%
  filter(reject) %>%
  nrow

n_Linda_16S_soil=
  Linda_16S_soil_df %>%
  filter(reject) %>%
  nrow

n_Linda_ITS_root_no_random=
  Linda_ITS_root_no_random_df %>%
  filter(reject) %>%
  nrow

n_Linda_ITS_soil_no_random=
  Linda_ITS_soil_no_random_df %>%
  filter(reject) %>%
  nrow

n_Linda_16S_root_no_random=
  Linda_16S_root_no_random_df %>%
  filter(reject) %>%
  nrow

n_Linda_16S_soil_no_random=
  Linda_16S_soil_no_random_df %>%
  filter(reject) %>%
  nrow


LINDA_summary=
  data.frame(Locus=c("ITS", "ITS", "16S", "16S"),
             SampleType=c("Root","Soil","Root","Soil"),
             n_LINDA_no_random=c(n_LINDA_ITS_root_no_random, n_LINDA_ITS_soil_no_random, n_LINDA_16S_root_no_random, n_LINDA_16S_soil_no_random),
             n_LINDA_random=c(n_LINDA_ITS_root, n_LINDA_ITS_soil, n_LINDA_16S_root, n_LINDA_16S_soil),
             n_ANCOM_no_random=c(n_ANCOM_ITS_root_no_random, n_ANCOM_ITS_soil_no_random, n_ANCOM_16S_root_no_random, n_ANCOM_16S_soil_no_random),
             n_ANCOM_random=c(n_ANCOM_ITS_root, n_ANCOM_ITS_soil, n_ANCOM_16S_root, n_ANCOM_16S_soil))

LINDA_summary
```
