---
title: "dbRDA_FinalScript"
output: html_document
date: "2025-01-09"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#The following script was used to generate dbRDA plots & accompanying statistics for the New Phytologist publication. The script is split into analysis by amplicon following initial data load and cleanup. 

#Set WD
```{r}
setwd("/Users/alex/Documents/Documents - Alex’s MacBook Pro/Documents/GitHub/Aus-Invasions-2023-Course/dbRDA")
```

#Load libararies
```{r}
library(tidyverse)
library(vegan)
library(dplyr)
library(readxl)
library(plyr)
```

#Reading in distance matrices & metadata
```{r}
#Includes all root & soil samples
#Community matrices by sample name
Bac = read.csv("/Users/alex/Documents/Documents - Alex’s MacBook Pro/Documents/GitHub/Aus-Invasions-2023-Course/dbRDA/OtuMat16S_pooled_rel.csv")
Fung = read.csv("/Users/alex/Documents/Documents - Alex’s MacBook Pro/Documents/GitHub/Aus-Invasions-2023-Course/dbRDA/OtuMatITS_pooled_rel.csv")
AMF = read.csv("/Users/alex/Documents/Documents - Alex’s MacBook Pro/Documents/GitHub/Aus-Invasions-2023-Course/dbRDA/OtuMat18S_pooled_rel.csv")

#Remove negative controls
Fung = Fung[-19,]
Bac = Bac[-19,]

#Removing root samples (since we are using soil env variables)
AMF = AMF[-c(1,3,5,7,9,11,13,15,17,19,21,23,25,27,29,31,33,35),]
Fung = Fung[-c(1,3,5,7,9,11,13,15,17,19,21,23,25,27,29,31,33,35),]
Bac = Bac[-c(1,3,5,7,9,11,13,15,17,19,21,23,25,27,29,31,33,35),]

#Read in metadata
Meta = read.csv("/Users/alex/Documents/Documents - Alex’s MacBook Pro/Documents/GitHub/Aus-Invasions-2023-Course/Merged_data/Aus23_allData_19Nov24.csv")
#Changing Eucalyptus to Eucalypt in metadata
Meta$TreeSpecies[Meta$TreeSpecies == "Eucalyptus"] <- "Eucalypt"
```

#Checking predictors for correlations (preemptive selection)
```{r}
#Subsetting env variables of interest
subset = Meta[,c(66,67,69,71,72,73,74,75,76,77,78,85)]

#Renaming variables to make the graphs more legible
subset = subset %>% 
  dplyr::rename(
    Soil_N = perc_N ,
    Soil_P = perc_P, 
    Soil_C = perc_C,
    Soil_Moisture = soil_moisture,
    Ergosterol = ergosterol,
    Soil_OM = organic_matter_LOI,
    Tree_DBH = dbh_cm,
    Tree_Height = height_m,
    Litter_Depth = litter_depth,
    Pine_Litter_Proportion = pine_litter_prop,
    Euc_Litter_Proportion =  euc_litter_prop,
    Litter_Biomass = total_litter_biomass
    )

#Checking correlations
library("Hmisc")
res2 <- rcorr(as.matrix(subset), type = "spearman")
res2
#Correlations based on p-values:
#Soil N -> Soil C, Ergosterol, Soil OM, Tree DBH, Pine Litter Prop
#Soil C -> Soil N, Ergosterol, Soil OM, Pine Litter Prop
#Ergosterol -> Soil N, Soil C, Euc Litter Prop
#Soil Moisture -> Litter depth, Pine Litter Prop, Euc Litter Prop
#Tree DBH -> Tree Height
#Tree Height -> Pine Litter Prop, Euc Litter Prop
#Pine Litter Prop -> Euc Litter Prop
###So I will be dropping, Soil C, Ergosterol, Soil OM, Tree Height, and Euc Litter Prop from the model
```

#Bacterial (16S) dbRDA with env predictors of interest
```{r}
#Moving column of sample names to row names
Bac2 <- Bac[,-1]
rownames(Bac2) <- Bac[,1]
#Formula (7 explanatory variables, may risk being overdetermined given small sample size)
dbrda_bac = capscale(formula = Bac2 ~ Soil_N + Soil_P + Soil_Moisture + Tree_DBH + Litter_Depth + Pine_Litter_Proportion + Litter_Biomass, data=subset, distance = "robust.aitchison", sqrt.dist=TRUE)
print(dbrda_bac) 

#Including these variables below lead to overfitting
#Litter_OLayer_ergosterol, Litter_OLayer_percP, Litter_OLayer_percN, Litter_OLayer_percC, Litter_avg_ergosterol
#Removed these for collinearity
#CN_ratio, Euc_litter_prop, Soil_OM, Soil_C, Ergosterol, Tree height

#Create summary
dbrda_bac_sum = summary(dbrda_bac)
print(dbrda_bac_sum)
dbrda_bac_sum$biplot

#ANOVA 
#Full model
anova(dbrda_bac)
#F = 1.0561, p = 0.009

#Testing axes
anova(dbrda_bac, by="axis")
#CAP1, F = 1.3098, p = 0.007

#Testing environmental variables
anova(dbrda_bac, by="terms")
#Soil_N (F = 1.2263, p = 0.001)
#Pine_Litter_Prop (F = 1.1120, p = 0.018)

#Plotting
plot(dbrda_bac)

#Pulling species scores
spec_scores = as.data.frame(scores(dbrda_bac)$species)
```

#Plotting bacterial dbRDA with subset of env predictors (significant predictors in bold, insig in black)
```{r}
### Plot a dbRDA
 smry <- summary(dbrda_bac)
 View(smry)
 df1  <- data.frame(smry$sites[,1:4])       #CAPS 1-4
 df1$Species<-Meta$TreeSpecies
 df1 <- df1[order(df1$Species),] 
 find_hull <- function(df1) df1[chull(df1$CAP1, df1$CAP2), ]
 hulls <- ddply(df1, "Species", find_hull)
 cent<-aggregate(cbind(df1$CAP1,df1$CAP2) ~ Species, data = df1, FUN = mean)
 segs<-merge(df1, setNames(cent, c('Species', 'V1','V2')), 
             by = 'Species', sort = TRUE)
 df1$seg1<-segs$V1
 df1$seg2<-segs$V2
 df2  <- data.frame(smry$biplot[,1:2])     # loadings for CAP1 & 2
 View(df2)
#Subsetting sig. vs. insignificant predictors (Soil N & Pine Litter)
 sig = df2[c(1,6),]
 insig = df2[c(2,3,4,5,7),]
 #Plot with no ellipses / hulls etc
 rda.plot <- ggplot(df1, aes(x=CAP1, y=CAP2, color = df1$Species)) + 
   geom_point(size=3) +
   scale_color_manual(values=c("#FF9900", "#000DCC"), name="Species",
                     breaks=c("Eucalypt", "Pine"),
                     labels=c("Eucalypt", "Pine")) +
   geom_hline(yintercept=0, linetype="dotted") +
   geom_vline(xintercept=0, linetype="dotted") +
   theme_classic()
 rda.plot 
  #Add env variables (black arrows are significant, grey arrows are insignificant)
 rda.biplot <- rda.plot +
   geom_segment(data=sig, aes(x=0, xend=CAP1, y=0, yend=CAP2), 
                color="black", arrow=arrow(length=unit(0.01,"npc"))) +
   geom_text(data=sig, 
             aes(x=CAP1,y=CAP2,label=rownames(sig),
                 hjust=0.5*(1-sign(CAP1)),vjust=0.5*(1-sign(CAP2))), 
             color="black", size=4,  fontface = "bold") +
  geom_segment(data=insig, aes(x=0, xend=CAP1, y=0, yend=CAP2), 
                color="black", arrow=arrow(length=unit(0.01,"npc"))) +
   geom_text(data=insig, 
             aes(x=CAP1,y=CAP2,label=rownames(insig),
                 hjust=0.5*(1-sign(CAP1)),vjust=0.5*(1-sign(CAP2))), 
             color="black", size=4) +
   xlab("dbRDA1 [7.53%]") +
   ylab("dbRDA2 [6.64%]") +
   ggtitle("Bacteria (16S)") +
   theme(plot.title = element_text(hjust = 0.5, size = 16, face="bold"),
         axis.text=element_text(size=14),
        axis.title=element_text(size=16),
        legend.text=element_text(size=15),
        legend.title=element_text(size=15))
rda.biplot 
```

#Fungal (ITS) dbRDA with env predictors of interest
```{r}
#Moving column of sample names to row names
Fung2 <- Fung[,-1]
rownames(Fung2) <- Fung[,1]
#Formula (7 explanatory variables, may risk being overdetermined given small sample size)
dbrda_fung = capscale(formula = Fung2 ~ Soil_N + Soil_P + Soil_Moisture + Tree_DBH + Litter_Depth + Pine_Litter_Proportion + Litter_Biomass, data=subset, distance = "robust.aitchison", sqrt.dist=TRUE)
print(dbrda_fung) 

#Including these variables below lead to overfitting
#Litter_OLayer_ergosterol, Litter_OLayer_percP, Litter_OLayer_percN, Litter_OLayer_percC, Litter_avg_ergosterol
#Removed these for collinearity
#CN_ratio, Euc_litter_prop, Soil_OM, Soil_C, Ergosterol, Tree height

#Create summary
dbrda_fung_sum = summary(dbrda_fung)
print(dbrda_fung_sum)
dbrda_fung_sum$biplot

#ANOVA 
#Full model
anova(dbrda_fung)
#F = 1.0356, p = 0.045

#Testing axes
anova(dbrda_fung, by="axis")
#CAP1, F = 1.1626, p = 0.188

#Testing environmental variables
anova(dbrda_fung, by="terms")
#Soil_N (F = 1.8804, p = 0.013)
#Pine_Litter_Prop (F = 1.1308, p = 0.002)

#Plotting
plot(dbrda_fung)

#Pulling species scores
spec_scores = as.data.frame(scores(dbrda_fung)$species)
```

#Plotting fungal dbRDA with subset of env predictors (significant predictors in bold, insig in black)
```{r}
### Plot a dbRDA
 smry <- summary(dbrda_fung)
 View(smry)
 df1  <- data.frame(smry$sites[,1:4])       #CAPS 1-4
 df1$Species<-Meta$TreeSpecies
 df1 <- df1[order(df1$Species),] 
 find_hull <- function(df1) df1[chull(df1$CAP1, df1$CAP2), ]
 hulls <- ddply(df1, "Species", find_hull)
 cent<-aggregate(cbind(df1$CAP1,df1$CAP2) ~ Species, data = df1, FUN = mean)
 segs<-merge(df1, setNames(cent, c('Species', 'V1','V2')), 
             by = 'Species', sort = TRUE)
 df1$seg1<-segs$V1
 df1$seg2<-segs$V2
 df2  <- data.frame(smry$biplot[,1:2])     # loadings for CAP1 & 2
 View(df2)
#Subsetting sig. vs. insignificant predictors (Soil N & Pine Litter)
 sig = df2[c(1,6),]
 insig = df2[c(2,3,4,5,7),]
 #Plot with no ellipses / hulls etc
 rda.plot <- ggplot(df1, aes(x=CAP1, y=CAP2, color = df1$Species)) + 
   geom_point(size=3) +
   scale_color_manual(values=c("#FF9900", "#000DCC"), name="Species",
                     breaks=c("Eucalypt", "Pine"),
                     labels=c("Eucalypt", "Pine")) +
   geom_hline(yintercept=0, linetype="dotted") +
   geom_vline(xintercept=0, linetype="dotted") +
   theme_classic()
 rda.plot 
  #Add env variables (black arrows are significant, grey arrows are insignificant)
 rda.biplot <- rda.plot +
   geom_segment(data=sig, aes(x=0, xend=CAP1, y=0, yend=CAP2), 
                color="black", arrow=arrow(length=unit(0.01,"npc"))) +
   geom_text(data=sig, 
             aes(x=CAP1,y=CAP2,label=rownames(sig),
                 hjust=0.5*(1-sign(CAP1)),vjust=0.5*(1-sign(CAP2))), 
             color="black", size=4,  fontface = "bold") +
  geom_segment(data=insig, aes(x=0, xend=CAP1, y=0, yend=CAP2), 
                color="black", arrow=arrow(length=unit(0.01,"npc"))) +
   geom_text(data=insig, 
             aes(x=CAP1,y=CAP2,label=rownames(insig),
                 hjust=0.5*(1-sign(CAP1)),vjust=0.5*(1-sign(CAP2))), 
             color="black", size=4) +
   xlab("dbRDA1 [6.74%]") +
   ylab("dbRDA2 [6.54%]") +
   ggtitle("Fungi (ITS)") +
   theme(plot.title = element_text(hjust = 0.5, size = 16, face="bold"),
         axis.text=element_text(size=14),
        axis.title=element_text(size=16),
        legend.text=element_text(size=15),
        legend.title=element_text(size=15))
rda.biplot 
```

#AMF (18S) dbRDA with env predictors of interest
```{r}
#Moving column of sample names to row names
AMF2 <- AMF[,-1]
rownames(AMF2) <- AMF[,1]
#Formula (7 explanatory variables, may risk being overdetermined given small sample size)
dbrda_amf = capscale(formula = AMF2 ~ Soil_N + Soil_P + Soil_Moisture + Tree_DBH + Litter_Depth + Pine_Litter_Proportion + Litter_Biomass, data=subset, distance = "robust.aitchison", sqrt.dist=TRUE)
print(dbrda_amf) 

#Including these variables below lead to overfitting
#Litter_OLayer_ergosterol, Litter_OLayer_percP, Litter_OLayer_percN, Litter_OLayer_percC, Litter_avg_ergosterol
#Removed these for collinearity
#CN_ratio, Euc_litter_prop, Soil_OM, Soil_C, Ergosterol, Tree height

#Create summary
dbrda_amf_sum = summary(dbrda_amf)
print(dbrda_amf_sum)
dbrda_amf_sum$biplot

#ANOVA 
#Full model
anova(dbrda_amf)
#F = 1.0267, p = 0.145

#Testing axes
anova(dbrda_amf, by="axis")
#CAP1, F = 1.2863, p. = 0.004

#Testing environmental variables
anova(dbrda_amf, by="terms")
#Soil_N (F = 1.1806, p = 0.002)
#Tree DBH (F = 1.0785, p = 0.045)

#Plotting
plot(dbrda_amf)

#Pulling species scores
spec_scores = as.data.frame(scores(dbrda_amf)$species)
```

#Plotting amf dbRDA with subset of env predictors (significant predictors in bold, insig in black)
```{r}
### Plot a dbRDA
 smry <- summary(dbrda_amf)
 View(smry)
 df1  <- data.frame(smry$sites[,1:4])       #CAPS 1-4
 df1$Species<-Meta$TreeSpecies
 df1 <- df1[order(df1$Species),] 
 find_hull <- function(df1) df1[chull(df1$CAP1, df1$CAP2), ]
 hulls <- ddply(df1, "Species", find_hull)
 cent<-aggregate(cbind(df1$CAP1,df1$CAP2) ~ Species, data = df1, FUN = mean)
 segs<-merge(df1, setNames(cent, c('Species', 'V1','V2')), 
             by = 'Species', sort = TRUE)
 df1$seg1<-segs$V1
 df1$seg2<-segs$V2
 df2  <- data.frame(smry$biplot[,1:2])     # loadings for CAP1 & 2
 View(df2)
#Subsetting sig. vs. insignificant predictors (Soil N & Tree DBH)
 sig = df2[c(1,4),]
 insig = df2[c(2,3,5,6,7),]
#Renaming soil moisture because it goes off of plot
 row.names(insig)[2] = "SM"
 #Plot with no ellipses / hulls etc
 rda.plot <- ggplot(df1, aes(x=CAP1, y=CAP2, color = df1$Species)) + 
   geom_point(size=3) +
   scale_color_manual(values=c("#FF9900", "#000DCC"), name="Species",
                     breaks=c("Eucalypt", "Pine"),
                     labels=c("Eucalypt", "Pine")) +
   geom_hline(yintercept=0, linetype="dotted") +
   geom_vline(xintercept=0, linetype="dotted") +
   theme_classic()
 rda.plot 
  #Add env variables (black arrows are significant, grey arrows are insignificant)
 rda.biplot <- rda.plot +
   geom_segment(data=sig, aes(x=0, xend=CAP1, y=0, yend=CAP2), 
                color="black", arrow=arrow(length=unit(0.01,"npc"))) +
   geom_text(data=sig, 
             aes(x=CAP1,y=CAP2,label=rownames(sig),
                 hjust=0.5*(1-sign(CAP1)),vjust=0.5*(1-sign(CAP2))), 
             color="black", size=4,  fontface = "bold") +
  geom_segment(data=insig, aes(x=0, xend=CAP1, y=0, yend=CAP2), 
                color="black", arrow=arrow(length=unit(0.01,"npc"))) +
   geom_text(data=insig, 
             aes(x=CAP1,y=CAP2,label=rownames(insig),
                 hjust=0.5*(1-sign(CAP1)),vjust=0.5*(1-sign(CAP2))), 
             color="black", size=4) +
      xlab("dbRDA1 [7.48%]") +
   ylab("dbRDA2 [6.67%]") +
   ggtitle("AMF (18S)") +
   theme(plot.title = element_text(hjust = 0.5, size = 16, face="bold"),
         axis.text=element_text(size=14),
        axis.title=element_text(size=16),
        legend.text=element_text(size=15),
        legend.title=element_text(size=15))
rda.biplot 
```
